---
title: "Data Cleaning Report for Survey on Potato Production in Plateau State"
author: "Plateau State Potato Value Chain Support Project (PS-PVCP)"
date: "September 9, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## 1.0  INTRODUCTION

### 1.1 BACKGROUND 
The Government of the Federal Republic of Nigeria has received a loan from the African Development Fund in the amount of UA 8.0 million (about USD 11.2 million), to finance the Plateau State Potato Value Chain Support Project (PVCSP) which was to be implemented in 3 years (2018 – 2020) but has been extended twice, now to end by June 30, 2024. The project intervention is to contribute to employment generation; improve the competitiveness of the potato commodity value chain; and increase incomes of the commodity value chain actors. The project has the potential to create 60,000 jobs in the state of which 50% are to be women. 

The overall goal of the Potato Value Chain Support Project is to contribute to employment generation and shared wealth creation along the potato commodity value chain, as well as food security through increased access to rural infrastructure. The project’s specific objective is to increase, on a sustainable basis, the income of smallholder farmers and rural entrepreneurs that are engaged in the production, processing, storage and marketing of Potato. Key Indicators of outcomes of the Potato Value Chain Support Project include: additional income of farmers and rural entrepreneurs (50% are women), increased access to rural infrastructure works in the project area; and an improvement in the food security situation of food secure beneficiaries in the Project area (50% are women). An impact indicator is the additional jobs created (of which 50% are women). The total number of direct household beneficiaries targeted by the project is estimated at 11,500 households creating 60,000 jobs along the value chain of priority commodity of which 50% are women.

### 1.2	POTATO PRODUCTION IN PLATEAU STATE
Plateau State accounts for about 95% of the potato produced in Nigeria. Potato production in the State is done mainly during the rainy season and mostly in one cycle as farmers plant their crops early enough to avoid the devastation caused by Potato Late Blight. Also, prior to the year 2020, potato was only produced in nine (9) Local Government Areas (LGAs) of the State namely: Barkin Ladi, Bassa, Bokkos, Jos East, Jos North, Jos South, Mangu, Pankshin and Riyom. These nine (9) LGAs are referred to as Core Potato Producing LGAs while the remaining eight (8) LGAs are termed Non-Core Potato Producing LGAs and include: Kanam, Kanke, Mikang, Langtang North, Langtang South, Qua’an Pan, Shendam and Wase. These eight (8) Local Government Areas make up the Southern Agro-Ecological Zone of the State.

The Project Implementation Unit (PIU) of the Plateau State Potato Value Chain Support Project observed that climatic conditions of the Southern Agro-Ecological Zone of the State are similar to that of Zaria, Kaduna State where potato is produced as well through dry season irrigation. The PIU in the year 2020, conducted Potato Adaptability Trial in the Southern Agro-Ecological Zone on farmers’ farms from November, 2020 to February, 2021. The aim of the adaptability trial is to demonstrate the viability or otherwise of potato in that zone. The results of the trials were impressive and farmers were encouraged to key into potato production. This trial was revalidated in the year 2021 (November, 2021 to February, 2022) reaffirming the results obtained in the previous year.

With several workshops and trainings conducted for farmers across the State on potato commodity by the PIU, this survey seeks to provide an insight into farming practices relating to potato in the State as at the 2023 cropping season.



## 2.0  DATA COLLECTION
The survey process began in the first week of August, 2023 with the design of digital questionnaire using KoboToolbox. The choice of using KoboToolbox was based on its ability to function both offline and online bearing in mind the fact that potato farmers on the Plateau also reside in rural areas too, with some having poor or no network coverage. 

The KoboToolbox link was shared with the seventeen (17) Local Government Desk Officers on August 14, 2023 via the Whatsapp platform which has been in use since October, 2022. The survey ended on September 5, 2023, covering a duration of three weeks. The Plateau State Potato Value Chain Support Project has so far registered 24,227 farmers across the 17 Local Government Areas (LGAs) of the State through the various Desk Officers. 

About 26% of the registered farmers was targeted as sample size for the survey corresponding to 6,358 participants. Stratified Sampling was used to determine the sample size for each Local Government since the number of registered farmers vary across the LGAs. At the end of the survey, 4,747 farmers participated being 19.6% of the total registered farmers and 74% of the target sample size. The survey data (in csv format) is stored in GitHub, an open source developer platform and can be accessed via [this link](https://github.com/spomaria/Potato-Production-Survey-Sept-2023).

## 3.0  DATA CLEANING
Just as in most developing countries of the world, the locals use measurements that vary from international standards. Bearing this in mind, the survey questions were designed in a way as to use measurements that are familiar to the farmers and afterwards, make the necessary conversions. This technique will enhance farmers’ understanding of the survey questions and guarantee accurate responses thereby improving on the reliability of the results from survey.

Data Cleaning measures undertaken include:

1.	Farmers were requested to provide their phone numbers or those of a close relative or friend. Part of the phone number and other socio-economic characteristics of the farmers such as gender and year of birth, were combined together to generate unique identification code for each farmer. This unique code will be used for subsequent tracking in future surveys. Farmers’ phone numbers were then expunged from the data set to protect the privacy of those who partook in the survey and shield them from any sort of possible retribution or undesirable intrusion.

2.	From field experiences, it was gathered that farmers either purchased potato seeds in baskets measuring about 25kg or bags measuring about 50kg. The farmers were asked the number of baskets or bags of potato seeds bought. This was later converted to tonnes (ton) by multiplying the number of baskets by 25 and the number of bags by 50, summing and dividing by 1000. This gives the estimated total amount of potato seeds planted (ton) per farmer.

3.	Farmers were asked the number of plots of farmland cultivated not hectares. This is because, land is measured in plots in the State with a plot measuring 50 feet by 100 feet (15m × 30m). The number of plots was then multiplied by 450 and then divided by 10,000 thereby, converting the size of farmland to hectares.

4.	The data was cleaned by removal of unnecessary spaces in answers that were self-typed by respondents, filling in blank spaces with values such as zero which will not skew the results but will not hinder the software from giving accurate results either.

5. For the purpose of analysis, responses of farmers from the non-core Potato Producing LGAs were filtered out. This is because, the climatic conditions in the Southern Agro-ecological Belt of the State can only support dry season production of potato. Since this survey covers rainy season, only the core Potato Producing LGAs can realistically grow potato.

6. Responses from farmers whose harvest month and survey month disagree were also filtered out (e.g. farmer filled the questionnaire in August but claims to have harvested potato in September).

7. Finally, responses containing outliers where filtered out of the data. The data used for analysis was trimmed down to 3000 responses.

```{r loading_packages, results='hide'}
  # We now load the required packages for the analysis
  library(tidyverse)
  library(skimr)
```

```{r results='hide'}
getwd()
```

```{r quick_scheme, results='hide'}
# The name of the survey dataset is 'ps-pvcp_survey_july_2023--cleaned.csv'
# This data set is in the same folder as this script
pvcp_survey_data <- read_csv("/home/spomaria/Documents/Spomary-Files/R-Scripts/Potato-Production-Survey-Sept-2023/ps-pvcp_survey_sept_2023.csv")
# The next line of code gives the descriptive properties used in the narrative below
skim_without_charts(pvcp_survey_data)
```

```{r creation_of_variables, results='hide'}
#We now create variables relevant to the objectives 
#of our research
#Create a variable that categorizes the LGAs into 
#Core and Non-Core Potato Producing LGAs
pvcp_data <- pvcp_survey_data %>%
  #We use the pipe operator '%>%' to pass the result obtained from the 
  #previous line to the next line
  mutate(core_potato_lga = ifelse(lga_of_residence == "Barkin Ladi"|
                                    lga_of_residence == "Bassa"|
                                    lga_of_residence == "Bokkos"|
                                    lga_of_residence == "Jos East"|
                                    lga_of_residence == "Jos North"| 
                                    lga_of_residence == "Jos South"|
                                    lga_of_residence == "Riyom"| 
                                    lga_of_residence == "Mangu"|
                                    lga_of_residence == "Pankshin", "Yes", "No"),
         #Create another variable for the proportion of money spent on
         #preparation of potato farm compared to other crops
         farm_prep_prop = round(cost_of_farm_prep/
                                  (cost_of_farm_prep +
                                      farm_prep_cost_other_crops)*100, 2),
         #Create another variable for the proportion of fertilizer bought
         #compared to the total fertilizer needed
         fert_prop = round(fert_bags_purchased_preseason/fert_bags_required *100, 2),
         #variable representing time to complete the survey
         time_taken = end - start,
         #create a variable for day of the week
         survey_day = weekdays(start),
         #create a variable for month the questionnaire was completed by the farmer
         survey_month = months(start),
         #convert farm size from plot to hectare
         potato_farmsize_ha = farmland_cultivated_ha*450/10000,
         # convert seeds from bags and baskets to tons and round to 2 decimal places
         potato_seeds_tons = round((bags_of_potato_seeds*50 + baskets_of_potato_seed*25)/1000,2), 
         # convert farm size from plots to hectares and round to 2 decimal places
         farmsize_others_ha = round(farmland_other_crops*450/10000,2), 
         # convert potato quantity from bags to tons and round to 2 decimal places
         potato_harvested_tons = round(quantity_harvested_bags*50/1000,2), 
         qty_potato_stored_tons = round(bags_of_potato_stored*50/1000,2),
         post_harvest_loss_tons = round(post_harvest_loss_bags*50/1000,2),
         potato_sold_tons = round(potato_sold_bags*50/1000,2),
         potato_price_kg = round(potato_sold_bags/50,2),
         potato_price_ton = round(potato_sold_bags/50*1000,2))

#let's view our new dataset
skim_without_charts(pvcp_data)
```

```{r dirty_data}
pvcp_data %>%
  select(-c(phone_number, phone_number_yours, yahoo_address, gmail_address)) %>%
  write_csv(., 
            "/home/spomaria/Documents/Spomary-Files/R-Scripts/Potato-Production-Survey-Sept-2023/ps_pvcp_production_data_sept_2023--dirty.csv")
```
Saving the data in its raw form without before cleaning. This is useful for future data cleaning exercises.


```{r count_by_weekday, results='hide'}
#To find the busiest day of the survey
# We arrange the counts by day in descending order
pvcp_data %>% 
  #this code creates a count of responses per weekday
  count(survey_day) %>%
  # This code arranges the data in descending order
  # Note '-n' means in descending order
  arrange(-n)
```
```{r count_by_month, results='hide'}
#To find the busiest month of the survey
# We arrange the counts by day in descending order
pvcp_data %>% 
  #this code creates a count of responses per weekday
  count(survey_month) %>%
  # This code arranges the data in descending order
  # Note '-n' means in descending order
  arrange(-n)
```

```{r summary, results='hide'}
# We now clean the data to make our result reliable

pvcp_data %>%
  # We shall only use responses from the core Potato Producing LGAs for our analysis
  filter(core_potato_lga == "Yes") %>%
  filter(Household_size <100) %>%
  filter(farmland_cultivated_ha <100) %>%
  # Potato was introduced in Nigeria in the 1920's
  filter(years_of_experience <60) %>%
  # The normal price range for a bag of potato
  filter(potato_bag_price >15000 & potato_bag_price < 40000) %>%
  filter(baskets_of_potato_seed <1000) %>%
  # This is to to avoid having infinity as proportion of fertilizer purchased
  filter(fert_bags_required >1) %>%
  # Ensure there is coherence between the quantity of fertilizer purchased
  # and the estimated required quantity of fertilizer per farmer
  filter(fert_bags_purchased_preseason < fert_bags_required*2) %>%
  # Ensure there is coherence between quantity of potato harvested vs stored
  filter(qty_potato_stored_tons < potato_harvested_tons) %>%
  # Ensure consistency between quantity of potato harvested and size of farmland cultivated
  filter(potato_harvested_tons < farmland_cultivated_ha*50) %>%
  # Ensure consistency between quantity of potato harvested and quantity sold
  filter(potato_sold_bags < quantity_harvested_bags) %>%
  # We now glance through the summary statistics to check for unrealistic results
  summary()
```

```{r results='hide'}
colnames(pvcp_data)
```

```{r cleaned_data}
pvcp_data %>%
  # Select the desired attributes for our data analysis
  select(id, today, survey_day, survey_month, lga_of_residence, Village, time_taken, birth_year, 
         Gender, Marital_Status,
         Household_size, Educational_Level, Living_Arrangement, roof_type, coop_society_member,
         planting_time, potato_farmsize_ha, potato_seeds_tons, cost_of_seeds, cost_of_farm_prep,
         total_fert_bags_purchased, total_cost_of_fert, mixed_cropping, years_of_experience,
         prior_participation, potato_variety, time_of_harvest, potato_harvested_tons, need_a_store,
         find_storage_facility, qty_potato_stored_tons, post_harvest_losses, post_harvest_loss_tons,
         potato_sold_tons, potato_price_kg, potato_price_ton, hired_labourers, hired_labourers_num,
         labourers_wages, nature_of_labour) %>%
  # Change all column names (variable names) to lower case
  rename_all(., .funs = tolower) %>%
  # Remove any row that contains an empty cell
  drop_na() %>%
  # Save the file in a desired location for future use
  write_csv(., 
            "/home/spomaria/Documents/Spomary-Files/R-Scripts/Potato-Production-Survey-Sept-2023/ps_pvcp_production_data_sept_2023.csv")
```

This concludes the data cleaning exercise for this survey. Remember, the data both dirty and clean (in csv format) can be accessed via [this link](https://github.com/spomaria/Potato-Production-Survey-Sept-2023).
